Probabilistic Inference of Basic Stellar Parameters:\\ Application to Flickering Stars \altaffilmark{\titledag}

### 2016-04-01 version1.0/manuscript.tex
final changes

The two models we use to describe the relationships between $F_8$, \logg\ and
\rhostar\ are
\begin{equation}
% 	\log_{10}(\rho_\star) \sim \mathcal{N} \left(\mu = \alpha + \beta
% 	\log_{10}(F_8), \sigma = \sqrt{\sigma_{\rho}^2 + \gamma F_8}\right),
	\log_{10}(F_8) \sim \mathcal{N} \left(\mu = \alpha_\rho +
    \beta_\rho \log_{10}(\rho_\star), \sigma^2 + \sigma_\rho^2 \right),
\end{equation}
\label{eq:rho}
and
\begin{equation}
% 	\log_{10}(g) \sim \mathcal{N}\left(\mu = \delta + \epsilon \log_{10}(F_8),
% 	\sigma = \sqrt{\sigma_g^2 + \zeta F_8}\right).
	\log_{10}(F_8) \sim \mathcal{N} \left(\mu = \alpha_g + \beta_g
    \log_{10}(g), \sigma^2 + \sigma_g^2 \right).
\end{equation}
\label{eq:logg}
The free parameters of the two models are $\alpha_\rho$, $\beta_\rho$,
$\sigma_\rho$, $\alpha_g$, $\beta_g$ and $\sigma_g$.
These relations are Gaussian distributions with means given by the equation of
a straight line, and standard deviations which describe the intrinsic scatter
about the mean.
We used the MCMC package, {\tt emcee} \citep{foreman-mackey:2013} to explore the
posterior PDFs of our model parameters.

We also tested a model in which the additional scatter depends on flicker
itself, defined as
\begin{equation}
	\log_{10}(F_8) \sim \mathcal{N} \left(\mu = \alpha_\rho +
    \beta_\rho \log_{10}(\rho_\star), \sigma^2 + \sigma_{\rho}^2 +
    \gamma_\rho F_8 \right),
\end{equation}
\label{eq:rho2}
for flicker vs $\rho_\star$ and similarly for \logg.
This model allowed us to determine whether there the magnitude of additional
scatter varied as a function of flicker.
In other words, whether flicker was a better proxy for \logg\ or \rhostar\ for
either dwarf or giant stars.
We found that the maximum {\it a-posteriori} values for the $\gamma$
parameters were consistent with zero: $\gamma_{rho} = 0.006 \pm 0.02, \gamma_g
= -0.01 \pm\ 0.01$, and interpret this as evidence for a constant intrinsic
scatter level across evolutionary stages.

@@

We used a likelihood function which accounts for 2-D uncertainties but does
not allow the intrinsic scatter to be a function of the dependent or
independent variables.
For the relation between flicker and \rhostar, this likelihood function can be
written as
\begin{eqnarray}
	& p(F_8|\rho_\star, \alpha_\rho, \beta_\rho, \sigma_{\rho}) \propto  \\ \nonumber
						      & \exp \left[-\frac{1}{2}
		\sum_{n=1}^N \frac{[F_{8n}-(\alpha_\rho + \beta_\rho \rho_{*n})]^2}
	{\left[\beta_\rho \sigma_{F8, n}^2 + \sigma_{\rho *, n}^2
	+ \sigma_{\rho}^2\right]}\right]
	\\ \nonumber
	& + \beta_\rho^2 \sigma_{F8, n}^2 + \sigma_{\rho *, n}^2 + \sigma_{\rho}^2,
\end{eqnarray}
\label{eq:likelihood}
and similarly for \logg.
We found that the posterior PDFs for the model parameters obtained using this
likelihood function were consistent with those obtained using a model that
only accounts for the uncertainties on the flicker measurements.
The median values of the model parameters differed by around $0.05 \sigma$
for the $\alpha$ and $\beta$ parameters, by $0.3 \sigma$ for $\sigma_\rho$
and by 0.8 $\sigma$ for $\sigma_g$.
Since they are so dependent on the observational uncertainties, the parameters
that describe the intrinsic scatter in the relations are more sensitive to
whether the uncertainties in the $x$-direction are included.
Accounting for uncertainties on $y$ {\it and} $x$ is not essential in this
case but is still good practise and will, in general, produce more accurate
model parameters and uncertainties.

@@

We also tested uniform, flat priors as defined below:
\begin{eqnarray}
 	&	\alpha, \beta \sim U(-10:10) \\ \nonumber
  	&	\log (\sigma_{rho}), \log(\sigma_g) \sim U(-10:10).
 \end{eqnarray}
 \label{eq:priors}
We found that the results were relatively insensitive to the choice of prior,
with median parameter values differing by only around $0.05\sigma$.
MCMC chains were run until the Gelman \& Rubin convergence criterion,
$\hat{R}$ reached a value of less than 1.002 and the number of autocorrelation
times was greater than 35.
Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval which are
representative of the intrinsic scatter in the relations.
The marginal posterior PDFs of the model parameters for \rhostar are shown in
figure \ref{fig:triangle}.
The marginal posterior PDFs for \logg are similarly Gaussian and, as with
$\sigma_\rho$, $\sigma_g$ is clearly greater than zero.

@@

\begin{table}
\begin{center}
\caption{Median parameter values with 1$\sigma$ uncertainties.}
\begin{tabular}{lcc}
\hline\hline
Parameter & Median value \\
    \hline
$\alpha_\rho$   &    1.315$\pm 0.008$   \\
$\beta_\rho$    &    -0.526$\pm 0.009$   \\
$\sigma_\rho$   &    0.065$\pm 0.003$   \\
% 1.31476104613 0.00759733599155 0.00767814819025
% -0.525867117083 0.00913323668806 0.00926420817321
% 0.0650160215492 0.00300880167448 0.00288553857806
\hline
$\alpha_g$      &    4.90$\pm 0.05$   \\
$\beta_g$       &   -0.83$\pm 0.01$   \\
$\sigma_g$      &    0.056$\pm 0.003$   \\
% 4.90451472909 0.0514398889837 0.0509386027456
% -0.8252952869 0.0130730719222 0.0131097233196
% 0.0565263079576 0.00279179051697 0.00271277097572
    \hline
\end{tabular}
\end{center}
\end{table}
\label{tab:results}

@@

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity, whilst including parameters to describe
the intrinsic scatter in these relationships, presented in table
\ref{tab:results}.
The additive terms, $\sigma_\rho$ and $ \sigma_g$ are both non-zero,
suggesting that there {\it is} an additional source of scatter in the
relations, not accounted for by the observational uncertainties alone.
This is either caused by intrinsic scatter in the physical relationship
between flicker and density and \logg, produced by some physical process that
is not accounted for in the model, or by an underestimation of the
observational uncertainties.
We also tested a model with both an additive variance term {\it and} a term
that included flicker-dependent variance.
We found that the need for additional flicker-dependent variance was not
supported by the data, indicating that the intrinsic scatter in the relations
between flicker, \logg\ and \rhostar\ does not depend on evolutionary state.

@@

We provide posterior PDF samples and the code used in this project at
\url{https://zenodo.org/deposit/95681/}.
% https://github.com/RuthAngus/flicker
Whenever a prediction for the surface gravity or density of a star is required,
for a given estimate of flicker, we recommend using these posterior samples
within the calculation of \rhostar or \logg\ and its (Monte Carlo)
uncertainty.
These posterior samples will naturally fold in the covariances between
parameters.
Simple analytical uncertainty propagation is only valid when uncertainties are
Gaussian and uncorrelated which is rarely true and certainly not the case when
the model is a straight line (the slope and intercept are alway correlated).
A flicker value with uncertainties (or even better: posterior PDF
samples), input into our model will result in a probability distribution over
stellar densities or surface gravities which reflects both the uncertainties
on the flicker measurement, the uncertainties on the model parameters {\it and}
the intrinsic scatter in the flicker-\rhostar-\logg\ relations.

@@

RA would like to thank Dan Foreman-Mackey for his extremely helpful suggestions
and comments, Angie Wolfgang for her expert guidance and John Johnson for his
fantastic advice and support.
RA would also like to thank the anonymous referee for their excellent
suggestions, SAMSI institute and the other members of the SAMSI {\it Kepler}
exoplanet statistics working groups.

@@

% #####################################################################
%% Bibliography
\begin{thebibliography}{99}
\bibitem[\protect\citeauthoryear{Andrieu \& Roberts}{2009}]{andrieu:2009}
Andrieu, C., \& Roberts, G.~O., Ann. Statist., 2, 695
\bibitem[\protect\citeauthoryear{Angus et al.}{2015}]{angus:2015}
Angusr, R., Aigrain, S., Foreman-Mackey, D. \& McQuillan, A., 2015, MNRAS,
450, 1787
\bibitem[\protect\citeauthoryear{Bastien et al.}{2013}]{bastien:2013} Bastien,
F. A., Stassun, K. G., Basri, G. \& Pepper, J., 2013, Nature, 500, 427
\bibitem[\protect\citeauthoryear{Chaplin et al.}{2014}]{chaplin:2014}
Chaplin, W.~J., Basu, S., Huber, D. et al., 2014, ApJS, 210, 1
\bibitem[\protect\citeauthoryear{Cranmer et al.}{2014}]{cranmer:2014}
Cranmer, S. R., Bastien, F. A., Stassun, K. G. \& Saar, S. H., 2014, ApJ, 781,
124
\bibitem[\protect\citeauthoryear{Foreman-Mackey et al.}{2014}]
{foreman-mackey:2014} Foreman-Mackey, D., Hogg, D.~W. \& Morton, T.~D., 2014,
ApJ, 795, 64
\bibitem[Foreman-Mackey et al.(2013)]{foreman-mackey:2013}
Foreman-Mackey, D., Hogg, D.~W., Lang, D., \& Goodman, J.\ 2013, \pasp, 125, 306
\bibitem[Foreman-Mackey et al. (2014)]{corner}
Foreman-Mackey, D., Price-Whelan, A., Ryan, G. et al.\ 2014,
10.5281/zenodo.10598, http://dx.doi.org/10.5281/zenodo.10598
\bibitem[\protect\citeauthoryear{Fox}{1997}]{fox:1997}
Fox, J., 1997, Applied Regression Analysis, Linear Models, and Related Methods
(Thousand Oaks:Sage Publications, Inc.)
\bibitem[\protect\citeauthoryear{Fuller}{1987}]{fuller:1987}
Fuller, W.~A., 1987, Wiley Series in Probability and Mathematical Statistics
\bibitem[\protect\citeauthoryear{Hogg et al.}{2010}]{hogg:2010}
Hogg, D.~W., Myers, A.~D. \& Bovy, J., 2010, ApJ, 725,
2166
\bibitem[\protect\citeauthoryear{Hogg et al.}{2010b}]{hogg:2010b}
Hogg, D.~W., Bovy, J. \& Lang, D., 2010
\bibitem[\protect\citeauthoryear{Huber et al.}{2013}]{huber:2013}
Huber, D., Chaplin, W.~J., Christensen-Dalsgaard, J. et al., 2013, ApJ, 767,
127
\bibitem[\protect\citeauthoryear{Kelly}{2007}]{kelly:2007} Kelly, B.~C., 2007,
ApJ, 665, 1489
\bibitem[\protect\citeauthoryear{Kipping et al.}{2014}]{kipping:2014}
Kipping, D.~M., Bastien, F.~A., Stassun, K.~G., Chaplin, W.~J., Huber, D. \&
Buchhave, L.~A., 2014, ApJ, 785, 32
\bibitem[\protect\citeauthoryear{Rogers}{2015}]{rogers:2015}
Rogers, L.~A., 2015, ApJ, 801, 41
\bibitem[\protect\citeauthoryear{Seager \& Mall\'{e}n-Ornelas}{2003}]
{seager:2003}Seager, S., \& Mall\'{e}n-Ornelas, G., 2003, ApJ, 585, 1038
\bibitem[VanderPlas(2014)]{vanderplas}
VanderPlas, J.\ 2014, arXiv:1411.5018
\bibitem[\protect\citeauthoryear{Wolfgang et al.}{2015}]{wolfgang:2015}
Wolfgang, A., Rogers, L.~A., Ford, E.~B., 2015, ApJ, submitted
(astro-ph:1504.07557)
\end{thebibliography}

>>>>

The two models we use to describe the relationships between $F_8$, \logg\ and
\rhostar\ are
\begin{equation}
	\log_{10}(F_8) \sim \mathcal{N} \left(\alpha_\rho +
    \beta_\rho \log_{10}(\rho_\star), \sigma_\rho^2 \right),
\end{equation}
\label{eq:rho}
and
\begin{equation}
	\log_{10}(F_8) \sim \mathcal{N} \left(\alpha_g + \beta_g
    \log_{10}(g), \sigma_g^2 \right).
\end{equation}
\label{eq:logg}
The free parameters of the two models are $\alpha_\rho$, $\beta_\rho$,
$\sigma_\rho$, $\alpha_g$, $\beta_g$ and $\sigma_g$.
These relations are Gaussian distributions with means given by the equation of
a straight line and standard deviations which describe the intrinsic scatter
about the mean.
We used the MCMC package, {\tt emcee} \citep{foreman-mackey:2013} to explore the
posterior PDFs of our model parameters.

We also tested a model in which the additional scatter depends on flicker
itself, defined as
\begin{equation}
	\log_{10}(F_8) \sim \mathcal{N} \left(\mu = \alpha_\rho +
    \beta_\rho \log_{10}(\rho_\star), \sigma_{\rho}^2 + \gamma_\rho
    \log_{10}(F_8)
    \right),
\end{equation}
\label{eq:rho2}
for flicker vs $\rho_\star$ and similarly for \logg.
This model allowed us to determine whether there the magnitude of additional
scatter varied as a function of flicker.
In other words, whether flicker was a better proxy for \logg\ or \rhostar\ for
either dwarf or giant stars.
We found that the maximum {\it a-posteriori} values for the $\gamma$
parameters were consistent with zero: $\gamma_{rho} = 0.006 \pm 0.02, \gamma_g
= -0.01 \pm\ 0.01$, and interpret this as evidence for a constant intrinsic
scatter level across evolutionary stages.

@@

We used a likelihood function which accounts for 2-D uncertainties but does
not allow the intrinsic scatter to be a function of the dependent or
independent variables.
For the relation between flicker and \rhostar, this likelihood function can be
written as
\begin{eqnarray}
	& \ln\left[p(F_8|\rho_\star, \alpha_\rho, \beta_\rho, \sigma_{\rho})\right]
        \propto  \\ \nonumber
    & -\frac{1}{2}\sum_{n=1}^N \left[\frac{[F_{8n}-(\alpha_\rho
    + \beta_\rho \rho_{*n})]^2}
	{\left[\beta_\rho \sigma_{F8, n}^2 + \sigma_{\rho *, n}^2 +
    \sigma_{\rho}^2\right]} + \ln(\sigma_{F8, n}^2) + \ln(\sigma_{\rho*, n}^2)
    + \ln(\sigma_\rho^2) \right]
	\\ \nonumber
\end{eqnarray}
\label{eq:likelihood}
and similarly for \logg.
We found that the posterior PDFs for the model parameters obtained using this
likelihood function were consistent with those obtained using a model that
only accounts for the uncertainties on the flicker measurements.
{\bf The median values of the model parameters differed by around $0.05
\sigma$ for the $\alpha$ and $\beta$ parameters, by $0.3 \sigma$ for
$\sigma_\rho$ and by $0.8 \sigma$ for $\sigma_g$.
Since they are so dependent on the observational uncertainties, the parameters
that describe the intrinsic scatter in the relations are more sensitive to
whether the uncertainties in the $x$-direction are included.}
Accounting for uncertainties on $y$ {\it and} $x$ is not essential in this
case but is still good practise and will, in general, produce more accurate
model parameters and uncertainties.

@@

We also tested uniform, flat priors as defined below:
\begin{eqnarray}
 	&	\alpha, \beta \sim U(-10:10) \\ \nonumber
  	&	\log (\sigma_{rho}), \log(\sigma_g) \sim U(-10:10).
 \end{eqnarray}
 \label{eq:priors}
We found that the results were relatively insensitive to the choice of prior,
{\bf with median parameter values differing by only around $0.05\sigma$.
MCMC chains were run until the Gelman \& Rubin convergence criterion,
$\hat{R}$ reached a value of less than 1.002 and the number of autocorrelation
times was greater than 35.}
Figures \ref{fig:rhostar} and \ref{fig:logg} show the data with the best-fit
models.
The shaded regions show the 1 and 2$\sigma$ confidence interval which are
representative of the intrinsic scatter in the relations.
{\bf The marginal posterior PDFs of the model parameters for \rhostar are
shown in figure \ref{fig:triangle}.
The marginal posterior PDFs for \logg\ are similarly Gaussian and, as with
$\sigma_\rho$, $\sigma_g$ is clearly greater than zero.
We checked the consistency between the two relations by calculating flicker
values for the Sun, finding $F_8 = 1.24 \pm 0.07$ and $F_8 = 1.21 \pm 0.1$
from Solar density and surface gravity measurements, respectively.
All the code used for this project and several ipython notebooks explaining
our analysis are available at {\url https://github.com/RuthAngus/flicker}.}

@@

\begin{table}
\begin{center}
\caption{Median parameter values with 1$\sigma$ uncertainties.}
\begin{tabular}{lcc}
\hline\hline
Parameter & Median value \\
    \hline
$\alpha_\rho$   &    1.31$\pm 0.01$   \\
$\beta_\rho$    &    -0.53$\pm 0.01$   \\
$\sigma_\rho$   &    0.060$\pm 0.003$   \\
\hline
$\alpha_g$      &    4.91$\pm 0.05$   \\
$\beta_g$       &   -0.83$\pm 0.01$   \\
$\sigma_g$      &    0.060$\pm 0.003$   \\
    \hline
\end{tabular}
\end{center}
\end{table}
\label{tab:results}

@@

We have recalibrated the relation between short timescale brightness
fluctuations in the {\it Kepler} light curves of stars (flicker) with both
stellar density and surface gravity, whilst including parameters to describe
the intrinsic scatter in these relationships, presented in table
\ref{tab:results}.
The terms $\sigma_\rho$ and $ \sigma_g$ are both non-zero, suggesting that
there {\it is} an additional source of scatter in the relations, not accounted
for by the observational uncertainties alone.
This is either caused by intrinsic scatter in the physical relationship
between flicker and density and \logg, produced by some physical process that
is not accounted for in the model, or by an underestimation of the
observational uncertainties.
We also tested a model with both an additional variance term {\it and} a term
that included flicker-dependent variance.
We found that the need for additional flicker-dependent variance was not
supported by the data, indicating that the intrinsic scatter in the relations
between flicker, \logg\ and \rhostar\ does not depend on evolutionary state.

@@

We provide posterior PDF samples at \url{https://zenodo.org/deposit/105051/}.
% https://zenodo.org/deposit/95681/}.
Whenever a prediction for the surface gravity or density of a star is required,
for a given estimate of flicker, we recommend using these posterior samples
within the calculation of \rhostar or \logg\ and its (Monte Carlo)
uncertainty.
These posterior samples will naturally fold in the covariances between
parameters.
Simple analytical uncertainty propagation is only valid when uncertainties are
Gaussian and uncorrelated which is rarely true and certainly not the case when
the model is a straight line (the slope and intercept are alway correlated).
A flicker value with uncertainties (or even better: posterior PDF
samples), input into our model will result in a probability distribution over
stellar densities or surface gravities which reflects both the uncertainties
on the flicker measurement, the uncertainties on the model parameters {\it and}
the intrinsic scatter in the flicker-\rhostar-\logg\ relations.

@@

RA would like to thank Dan Foreman-Mackey for his extremely helpful suggestions
and comments, Angie Wolfgang for her expert guidance and John Johnson for his
fantastic advice and support.
RA would also like to thank the anonymous referee for their patience and
excellent suggestions which dramatically improved this letter, the SAMSI
institute and the other members of the SAMSI {\it Kepler} exoplanet statistics
working groups.

@@

% #####################################################################
%% Bibliography
\begin{thebibliography}{99}
\bibitem[\protect\citeauthoryear{Andrieu \& Roberts}{2009}]{andrieu:2009}
Andrieu, C., \& Roberts, G.~O., Ann. Statist., 2, 695
\bibitem[\protect\citeauthoryear{Angus et al.}{2015}]{angus:2015}
Angusr, R., Aigrain, S., Foreman-Mackey, D. \& McQuillan, A., 2015, MNRAS,
450, 1787
\bibitem[\protect\citeauthoryear{Bastien et al.}{2013}]{bastien:2013} Bastien,
F. A., Stassun, K. G., Basri, G. \& Pepper, J., 2013, Nature, 500, 427
\bibitem[\protect\citeauthoryear{Chaplin et al.}{2014}]{chaplin:2014}
Chaplin, W.~J., Basu, S., Huber, D. et al., 2014, ApJS, 210, 1
\bibitem[\protect\citeauthoryear{Cranmer et al.}{2014}]{cranmer:2014}
Cranmer, S. R., Bastien, F. A., Stassun, K. G. \& Saar, S. H., 2014, ApJ, 781,
124
\bibitem[\protect\citeauthoryear{Foreman-Mackey et al.}{2014}]
{foreman-mackey:2014} Foreman-Mackey, D., Hogg, D.~W. \& Morton, T.~D., 2014,
ApJ, 795, 64
\bibitem[Foreman-Mackey et al.(2013)]{foreman-mackey:2013}
Foreman-Mackey, D., Hogg, D.~W., Lang, D., \& Goodman, J.\ 2013, \pasp, 125, 306
\bibitem[Foreman-Mackey et al. (2014)]{corner}
Foreman-Mackey, D., Price-Whelan, A., Ryan, G. et al.\ 2014,
10.5281/zenodo.10598, http://dx.doi.org/10.5281/zenodo.10598
\bibitem[\protect\citeauthoryear{Fox}{1997}]{fox:1997}
Fox, J., 1997, Applied Regression Analysis, Linear Models, and Related Methods
(Thousand Oaks:Sage Publications, Inc.)
\bibitem[\protect\citeauthoryear{Fuller}{1987}]{fuller:1987}
Fuller, W.~A., 1987, Wiley Series in Probability and Mathematical Statistics
% \bibitem[\protect\citeauthoryear{Hogg et al.}{2010}]{hogg:2010}
% Hogg, D.~W., Myers, A.~D. \& Bovy, J., 2010, ApJ, 725,
% 2166
\bibitem[\protect\citeauthoryear{Hogg et al.}{2010b}]{hogg:2010b}
Hogg, D.~W., Bovy, J. \& Lang, D., 2010
\bibitem[\protect\citeauthoryear{Huber et al.}{2013}]{huber:2013}
Huber, D., Chaplin, W.~J., Christensen-Dalsgaard, J. et al., 2013, ApJ, 767,
127
\bibitem[\protect\citeauthoryear{Kelly}{2007}]{kelly:2007} Kelly, B.~C., 2007,
ApJ, 665, 1489
\bibitem[\protect\citeauthoryear{Kipping et al.}{2014}]{kipping:2014}
Kipping, D.~M., Bastien, F.~A., Stassun, K.~G., Chaplin, W.~J., Huber, D. \&
Buchhave, L.~A., 2014, ApJ, 785, 32
\bibitem[\protect\citeauthoryear{Rogers}{2015}]{rogers:2015}
Rogers, L.~A., 2015, ApJ, 801, 41
\bibitem[\protect\citeauthoryear{Seager \& Mall\'{e}n-Ornelas}{2003}]
{seager:2003}Seager, S., \& Mall\'{e}n-Ornelas, G., 2003, ApJ, 585, 1038
\bibitem[VanderPlas(2014)]{vanderplas}
VanderPlas, J.\ 2014, arXiv:1411.5018
\bibitem[\protect\citeauthoryear{Wolfgang et al.}{2015}]{wolfgang:2015}
Wolfgang, A., Rogers, L.~A., Ford, E.~B., 2015, ApJ, submitted
(astro-ph:1504.07557)
\end{thebibliography}
